SUPPLEMENTARY INFORMATION: Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-15 HGmodelMANUSCRIPT/HGmodel_ACSsupplementary.tex
Separate main text and supplementary

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[An \textsf{achemso} demo]
  {Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{IR,NMR,UV}
\keywords{American Chemical Society, \LaTeX}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents. It is given here as some journals
%% require that it is printed as part of the abstract page. It will
%% be automatically moved as appropriate.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{tocentry}

Some journals require a graphical entry for the Table of Contents.
This should be laid out ``print ready'' so that the sizing of the
text is correct.

Inside the \texttt{tocentry} environment, the font used is Helvetica
8\,pt, as required by \emph{Journal of the American Chemical
Society}.

The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
permitted for  \emph{Journal of the American Chemical Society}
graphical table of content entries. The box will not resize if the
content is too big: instead it will overflow the edge of the box.

This box and the associated title will always be printed on a
separate page at the end of the document.

\end{tocentry}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Phospholipids are essential building blocks of biological membranes.
Despite of vast amount of very accurate experimental data, the atomistic resolution structures sampled by the glycerol backbone and choline headgroup
in phoshatidylcholine bilayers are not known. Atomistic resolution molecular dynamics simulations 
have the potential to resolve the structures, and to give an arrestingly intuitive interpretation of the experimental data---but only if the simulations
reproduce the data within experimental accuracy.
In the present work, we simulated phosphatidylcholine (PC) lipid bilayers with 13 different
atomistic models, and compared simulations with NMR experiments in terms of the highly structurally sensitive C--H bond vector order
parameters. Focusing on the glycerol backbone and choline headgroups, 
we showed that the order parameter comparison can be used to judge the atomistic resolution structural accuracy of the models.
Accurate models, in turn, allow molecular dynamics simulations to be used as an interpretation tool that translates these NMR data
into a dynamic three dimensional representation of biomolecules in biologically relevant conditions.
In addition to lipid bilayers in fully hydrated conditions, we reviewed previous experimental
data for dehydrated bilayers and cholesterol-containing bilayers, and interpreted them with simulations.
Although none of the existing models reached experimental accuracy, by critically comparing them we were able to distill relevant chemical information:
(1) increase of choline order parameters indicates the P--N vector tilting more parallel to the membrane,
and (2) cholesterol induces only minor changes to the PC (glycerol backbone) structure.
%None of the current models is not accurately enough to resolve the structure with experimental accuracy.
%However, closer inspection of three best performing models (CHARMM36, GAFFlipid and MacRog) suggest
%that improvements in the sampled dihedral angle distributions would potentilly lead to the model which
%would resolve the structure. Despite of the inaccuracy in the fully hydrated
%structures, the response to dehydration, i.e. the tilting of the P--N vector more parallel to membrane normal, 
%is qualitatively correct in all models. The CHARMM36 and MacRog models describe the interactions
%between lipids and cholesterol better than the Berger/H\"oltje model.
This work has been done as a fully open collaboration, using \url{nmrlipids.blogspot.fi} as a
communication platform; all the scientific contributions were made publicly on this blog.
During the open research process, the repository holding our simulation trajectories and files
(\url{https://zenodo.org/collection/user-nmrlipids}) has become the most extensive
publicly available collection of molecular dynamics simulation trajectories of lipid bilayers.
\end{abstract}

@@

\section{Introduction}

Phospholipids containing various polar headgroups and acyl chains are essential building blocks of 
biological membranes. Lamellar phospholipid bilayer structures have been widely studied with various experimental 
and theoretical techniques as a simple model for cellular membranes~\cite{lipowsky95,tieleman97,klauda08,edholm08,tieleman10,piggot12,rabinovich13,marsh13}. 
Phospholipid molecules are composed of hydrophobic acyl chains connected by a glycerol backbone to a hydrophilic headgroup;
see Fig.~\ref{POPCstructure} for the structure of 1-palmitoyl-2-oleoyl\-phosphatidyl\-choline (POPC).
The behaviour of the acyl chains in a lipid bilayer is relatively well understood~\cite{Israelachvili80,lipowsky95,tieleman97,klauda08,edholm08,tieleman10,marsh13}. 
The conformations sampled by the glycerol backbone and choline in a fluid bilayer are, however, not fully 
resolved as even the most accurate scattering and Nuclear Magnetic Resonance (NMR)
techniques give only a set of values that the structure has to fulfill, but
there is no unique way to derive the actual structure from them~\cite{seelig77b,skarjune79,Israelachvili80,jacobs80,davis83,strenk85,akutsu91,hong95b,hong96,semchyschyn04}.
Some structural details have been extracted from crystal structures, $^1$H NMR studies, and Raman spectroscopy~\cite{hauser80,hauser81,hauser81b,akutsu81b,pascher92,hauser88,marsh06},
but general consensus concerning the structures sampled in the fluid state has not been reached~\cite{seelig77b,skarjune79,Israelachvili80,jacobs80,davis83,strenk85,hauser88,akutsu91,hong95b,hong96,semchyschyn04,marsh06}. 
Importantly, the structural parameters for the glycerol backbone are similar for various biologically
relevant lipid species (phosphatidylcholine (PC), phosphatidylethanolamine (PE) and phosphatidylglycerol (PG)) 
in various environments~\cite{gally81}, and the structural parameters for the choline headgroup are similar in model membranes and
real cells (mouse fibroblast L-M cell)~\cite{scherer87}.
Thus, resolving the PC-lipid glycerol and choline structures would be 
useful for understanding a wide range of different biological membranes.

Classical atomistic molecular dynamics simulations have been widely used to study  
lipid bilayers~\cite{tieleman97,klauda08,edholm08,tieleman10,piggot12,rabinovich13}. As these models provide an atomistic
resolution description of the whole lipid molecule, they have the potential to solve the glycerol backbone and 
headgroup structures. The experimental C--H bond order parameters
(routinely compared between experiments and simulations for the acyl chains~\cite{tieleman97,klauda08,edholm08,tieleman10,piggot12})
are also known for the glycerol backbone 
(g$_1$, g$_2$, and g$_3$) and choline ($\alpha$ and $\beta$) segments (see Fig.~\ref{POPCstructure} for definitions) and are among the main parameters used in
attempts to derive lipid structures from experimental data~\cite{seelig77b,skarjune79,jacobs80,davis83,akutsu91,hong95b,semchyschyn04}.
Notably, the structures sampled in a simulation that reproduces these %and other experimental
parameters will automatically comprise an interpretation of the experiments. In other words, such simulations can be considered as an accurate atomistic resolution description of
the behavior of lipid molecules in a bilayer.

Only a few studies~\cite{shinoda97,hogberg08,castro08,klauda10,kapla12,dickson12,poger12,ferreira13,chowdhary13,maciejewski14}
have compared
the glycerol backbone and choline headgroup order parameters between simulations and experiments. 
The main reason probably is that the existing experimental data for the glycerol backbone
and choline headgroups are scattered over many publications and published in a format that is difficult to understand without some NMR expertise. 
In addition to the order parameters, dihedral angles for the glycerol backbone and headgroup estimated from experiments have sometimes been used to 
assess the quality of a simulation model~\cite{robinson94,essex94,kothekar96,hyvonen97,shinoda97,duong99}.

In this work, we first review the most relevant experimental data for the glycerol backbone and choline headgroup order parameters
in a phosphatidylcholine lipid bilayer. Then the available atomistic resolution lipid models are carefully compared to the 
experimental data. The comparison reveals that the CHARMM36~\cite{klauda10}, GAFFlipid~\cite{dickson12}, and MacRog~\cite{maciejewski14} models
have the most realistic glycerol backbone and choline structures. We also compare the glycerol backbone and choline 
structures between the most often used (Berger-based) lipid model~\cite{berger97} 
%\todo{If there are no objections, we will leave the naming convention of different Berger versions as it is now.}
and the best performing models, to demonstrate that by using the 
order parameters we can distinguish the more reasonable structures from the less reasonable ones. However, none of the current models 
is accurate enough to properly resolve the atomistic resolution structures.

In addition to fully hydrated single component lipid bilayers, the glycerol backbone and choline order parameters
have been measured under a large number of changing conditions: hydration level~\cite{bechinger91,ulrich94,dvinskikh05b}, cholesterol content~\cite{brown78,ferreira13},
ion concentration~\cite{brown77,akutsu81,altenbach84,roux90,roux91}, temperature~\cite{gally75}, charged lipid content~\cite{roux90,roux91}, charged surfactant content~\cite{scherer89}, 
drug molecule concentration~\cite{browning82,kelusky84,castro08}, and protein content~\cite{roux89,kuchinka89} (listing only the publications most relevant for this work and the pioneering studies).
Existence of these data allows the comparison of structural responses to varying conditions between simulations and experiments,
in other words, validation of the simulation models and interpretation of the original experiments. 
Here we demonstrate the power of this approach in understanding the behaviour of a bilayer as a function of hydration level and cholesterol content.
Choline headgroup order parameters as function of ion concentration, and their relation to the ion binding affinity, are discussed elsewhere~\cite{ionpaper}.

  \begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{../Fig/POPCstructure.pdf}

  \caption{\label{POPCstructure}
    Chemical structure of  1-palmitoyl-2-oleoylphosphatidylcholine (POPC).}

\end{figure}

\section{Methods}

\subsection{Open collaboration}
%\todo{Should we write more about this?}

This work has been done as a fully open collaboration, using the \url{nmrlipids.blogspot.fi} blog~\cite{nmrlipids}
as a communication platform.
Our approach is inspired by the Polymath project~\cite{gowers09}, however there are some essential differences. 
We started by publishing a manuscript~\cite{ollila13} discussing the glycerol backbone and choline structures 
in a Berger-based model (the most used molecular dynamics simulation model for lipid bilayers).
Simultaneously,  we presented an open invitation for further contributions and discussion on the blog.
All the scientific contributions were made publicly through the blog. Every contributor was offered coauthorship
according to the guidelines defined in the beginning of the project~\cite{creditsPOST};
the acceptance of the offer was based on authors' self-assesment of their scientific contribution.
These contributions are summarized in the Supplementary Information. %(the location depends on the journal probably).

Almost all simulation data, including input files for reproduction and trajectories for further analysis, are collected
on our CERN-hosted Zenodo file repository (\url{https://zenodo.org/collection/user-nmrlipids}).
Thus, in addition to the main topic of this manuscript, we present the most extensive publicly available collection of simulation trajectories
for lipid bilayers, opening up numerous possiblities for different analyses with much less effort than previously required.
Further information, such as scripts, figures, and manuscript text files, are available through our GitHub 
repository (\url{https://github.com/NMRlipids}). \todo{I will not assign a doi yet. We can wait for the journal decision.}.

\subsection{Order parameters from experiments}\label{expORDp}
The order parameter of a hydrocarbon C--H vector is defined as 
\begin{equation}\label{orderP}
S_{{\rm CH}}=\frac{1}{2}\langle 3 \cos^2 \theta-1 \rangle,
\end{equation} 
where the angle brackets denote an ensemble average over the sampled conformations, and $\theta$ is the angle between the C--H bond and the membrane normal.
The absolute values of order parameters can be measured by detecting quadrupolar splitting with $^2$H NMR~\cite{seelig77c} or by detecting dipolar 
splitting with $^1$H-$^{13}$C NMR~\cite{hong95a,gross97,dvinskikh05a,ferreira13}. The measurements are based on
different physical interactions and also the connection between order parameters and quadrupolar or dipolar splitting
are different. The absolute values of order parameters from the measured quadrupolar splitting $\Delta \nu_Q$ ($^2$H NMR) are calculated using 
the equation $|S_{{\rm CD}}|=\frac{4}{3} \frac{e^2qQ}{h} \Delta \nu_Q$, where the value for the static quadrupole
splitting constant is estimated from various experiments to be 170~kHz leading to a numerical relation $|S_{{\rm CD}}|=0.00784 \times \Delta \nu_Q$~\cite{seelig77c}. 
The absolute values of order parameters from the effective dipolar coupling $d_\mathrm{CH}$ ($^1$H-$^{13}$C NMR) are calculated using equation
$|S_{{\rm CH}}|=\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c} d_\mathrm{CH}$, where
values between 20.2--22.7 kHz are used for $\frac{4\pi\langle r_\mathrm{CH}^3 \rangle}{\hbar \mu_0 \gamma_h \gamma_c}$,
depending on the original authors~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
The effective dipolar coupling $d_\mathrm{CH}$ is related to the measured dipolar splitting $\Delta \nu_{CH}$ 
through scaling factor which depends on the pulse sequence used in the $^1$H-$^{13}$C NMR experiment~\cite{hong95a,gross97,dvinskikh05a,ferreira13}.
It is important to note that the order parameters measured with different techniques based on different physical interactions are in good agreement
with each other (see Results and Discussion), indicating very high quantitative accuracy of the measurements.
For a more detailed discussion see Ref.~\citenum{accuracyPOST}.

The absolute values of order parameters are accessible with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. 
However, only $^1$H-$^{13}$C NMR techniques allow also the measurement of the sign of the order parameter~\cite{hong95a,hong95b,gross97}. 
The measured sign is negative for almost all the carbons discussed in this work, except for $\alpha$ which is positive~\cite{hong95a,hong95b,gross97}. 
For more detailed discussion about the determination of the sign of the order parameters see Ref.~\citenum{signpost}.

For most CH$_2$ segments in fluid phosholipid bilayer the order parameters are equal for both hydrogens attached to the same carbon.
However, in some cases (e.g. g$_1$, g$_3$ and  C$_2$ carbon in the \textit{sn}-2 chain) the order parameters are not equal and this 
can be observed with both $^2$H NMR and $^1$H-$^{13}$C NMR techniques. In the present work we call the phenomena of unequal order parameters 
for hydrogens attached to the same carbon as {\it forking} to avoid confusion with dipolar and quadrupolar splitting in NMR terminology. 
Forking has been studied in detail with $^2$H NMR techniques by deuterating the R or S position in CH$_2$ segment, and
the studies show that the forking arises from differently sampled orientations of the two C--H bonds, not from two 
separate populations of lipid conformations~\cite{engel81,gally81}.

\subsection{Order parameters from simulations}
The order parameters from simulations were calculated directly using the definition as in Eq.~\ref{orderP}.
%$S_{{\rm CH}}=\frac{3}{2} \langle \cos^2 \theta-1 \rangle$. 
For the united atom models the hydrogen positions were generated 
in the trajectories post-simulationally using the positions of the heavy atoms and the known hydrocarbon geometries.
For the statistical error estimates, the time average of order parameters were first calculated separately
for each lipid in the system. Then it was assumed that different lipids are statistically independent 
entities (which should be the case in fluid phase) and the error of the mean for the average over individual 
lipids in the system was calculated and used as error bar for the order parameters.

It has been recently pointed out that the sampling of individual dihedral angles might be very
slow compared to the typical simulation timescales~\cite{vogel12}. On the other hand, another recent
study shows that the slowest rotational correlation functions of a C--H bond (g$_1$) reaches a plateau (S$_{CH}^2$)
after 200~ns in the Berger-POPC-07~\cite{ollila07a} model, and that the dynamics of this segment is significantly too slow in simulations
compared to the experiments~\cite{ferreira15}. In practise, less than 200~ns of simulation data is enough for the order parameter
calculation due to the average over different lipid molecules. In conclusion, if the sampling with typical simulation times
is not enough for the convergence of the order parameters, then the simulation models has significantly too slow dynamics.

\subsection{Simulated systems}
All simulations are ran with a standard setup for a planar lipid bilayer in zero tension and constant temperature
with periodic boundary conditions in all directions by using the GROMACS software package~\cite{hess08} 
(version numbers 4.5.X--4.6.X), LAMMPS~\cite{plimpton95}, MDynaMix~\cite{Lyubartsev00} or NAMD~\cite{phillips05}.
The number of molecules, simulation temperatures and the length of simulations of all the simulated systems 
are listed in Tables~\ref{systems},~\ref{systemsDEHYD} and~\ref{systemsCHOL}. Full simulation
details are given in the Supplementary Information (SI) or in the original publications in case the
data is used previously therein. Additionally, the files related to the simulations and the resulting trajectories are publicly
available for almost all systems in the Zenodo collection \url{https://zenodo.org/collection/user-nmrlipids}. 
The references pointing to simulation details and files are also listed in Tables~\ref{systems},~\ref{systemsDEHYD} and~\ref{systemsCHOL}.

\begin{sidewaystable}[]
\centering
\caption{Fully hydrated single component lipid bilayer systems simulated for Fig.~\ref{HGorderparameters}:
  1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC),
  %1,2-dioleoyl-sn-glycero-3-phosphocholine (DOPC),
  dilauroylphosphatidylcholine (DLPC),
  dipalmitoylphosphatidylcholine (DPPC), and
  1-palmitoyl-2-oleoylphosphatidylcholine (POPC).
The {\bf bolded} systems were used also for Fig.~\ref{HGorderparameters2}.
$^a$ Number of lipid molecules.
$^b$ Number of water molecules.
$^c$ Temperature.
$^d$ Total simulation time.
$^e$ Time used for analysis.
$^f$ Reference link for the downloadable simulation files;
     the data sets marked with $^*$ also include a part of the trajectory.
$^g$ Reference for the full simulation details;
     the original publication is cited if simulation data from previously published work has been directly used,
     for other systems the simulation details are given in the Supplementary Information.
$^h$ Magnitudes from Fig.~S4 of Klauda et al.~\cite{klauda10}, signs matched to our simulations.
$^i$ Magnitudes from Fig.~9 of Dickson et al.~\cite{dickson12}, signs matched to our simulations.
}\label{systems}
\begin{tabular}{l l r r r r r c c}
%\hline
Force field & lipid & $^a$N$_{\rm l}$   &  $^b$N$_{\rm w}$ &  $^c$T (K)  &  $^d$t$_{{\rm sim}}$(ns) &  $^e$t$_{{\rm anal}}$ (ns) &  $^f$Files  &  $^g$Details\\
\hline
Berger-DMPC-04~\cite{gurtovenko04} & DMPC & 128 & 5097 & 323 & 130 & 100 & [\citenum{dmpcFILES}]$^*$       & [\citenum{miettinen09}] \\
Berger-DPPC-98~\cite{marrink98}    & DPPC &  72 & 2864 & 323 & 60 & 30 & [\citenum{bergerDPPCfiles}]$^*$ & SI \\
{\bf
Berger-POPC-07}~\cite{ollila07a}   & POPC & 128 & 7290 & 298 & 270 & 240 & [\citenum{bergerFILESpopc}]$^*$ & [\citenum{ferreira15}] \\
CHARMM36~\cite{klauda10}  & DPPC &  72 & 2189 & 323 &  30 &  25 & [\citenum{charmmFILESdppc}]$^*$    & SI \\
CHARMM36~\cite{klauda10}  & DPPC &  72 & 2189 & 323 &  130  & - & -   & [\citenum{klauda10}]$^h$  \\
CHARMM36~\cite{klauda10}  & POPC &  72 & 2242 & 303 &  30 &  20 & [\citenum{charmm36filesSHORT}]$^*$ & SI \\
{\bf
CHARMM36}~\cite{klauda10} & POPC & 128 & 5120 & 303 & 200 & 100 & [\citenum{charmm36files}]$^*$      & SI \\
%CHARMM36\cite{klauda10}\todoi{Not included in the Fig. 2. Should we remove or include?}        & DOPC   & 128 &  5120    & 303 & 150 & 100  & ?\todoi{Permanent link in progress by Hubert Santuz}  & SI \\
%\todoi{Permanent link for DOPC CHARMM36 simulation by Hubert Santuz is not required now. However, it would be probably useful in the future.}
MacRog~\cite{kulig15b}  & POPC & 288 & 12600 & 310 & 100 &  80 & [\citenum{macrogFILES}]$^*$      & SI  \\
{\bf
MacRog}~\cite{kulig15b} & POPC & 128 &  6400 & 310 & 400 & 200 & [\citenum{macrogCHOLfiles}]$^*$  & SI  \\
MacRog~\cite{kulig15b}  & POPC & 288 & 14400 & 310 &  90 &  40 & [\citenum{macrogdehydFILES}]$^*$ & SI  \\
GAFFlipid~\cite{dickson12}   & DPPC &  72 & 2197 & 323 &  90 & 50 & [\citenum{GAFFlipidFILESdppc}]$^*$ & SI \\
GAFFlipid~\cite{dickson12}   & DPPC &  72 & 2167 & 323 & 250 &  250 & -   &  [\citenum{dickson12}]$^i$   \\
{\bf
GAFFlipid}~\cite{dickson12}  & POPC & 126 & 3948 & 303 & 137 & 32 & [\citenum{GAFFlipidFILES}]$^*$ & SI \\
Lipid14 \cite{dickson14}         & POPC  & 72 & 2234 & 303 & 100 & 50  & [\citenum{lipid14files}]$^*$ & SI \\
Poger \cite{poger10}             & DPPC  & 128 & 5841 & 323 & 2$\times$100 & 2$\times$50 & [\citenum{pogerFILESpme1,pogerFILESpme2}]$^*$ & SI \\
Slipids \cite{jambeck12}          & DPPC & 128 & 3840 & 323 & 150 & 100 & [\citenum{slipidsFILES}]$^*$ & SI \\
Slipids \cite{jambeck12b}          & POPC & 128 & 5120 & 303 & 200 & 150 & [\citenum{slipidsFILESpopc}]$^*$ & SI \\
Kukol \cite{kukol09}          & POPC   & 512 & 20564 & 298 & 50 & 30  & [\citenum{kukolFILES}]$^*$ & SI \\
Chiu \cite{chiu09}      & POPC  & 128 & 3552  & 298 & 56 & 50  & [\citenum{chiuFILES}]$^*$ & SI \\
H\"ogberg08 \cite{hogberg08}    & DMPC &  98 & 3840 & 303 &  75 & 50 &  [\citenum{hogbergDMPCfiles}]$^*$ & [\citenum{hogberg08}] \\
H\"ogberg08 \cite{rabinovich14} & POPC & 128 & 3840 & 303 & 100 & 80 &  [\citenum{hogbergPOPCfiles}]$^*$ & [\citenum{rabinovich14}]  \\
Ulmschneiders \cite{Ulmschneider09}    & POPC  & 128 & 3328 & 310 & 100 & 50 & [\citenum{ulmschneiderFILES}]$^*$ & SI \\
Tj\"ornhammar14 \cite{tjornhammar14}   & DPPC  & 144 & 7056 & 323 & 200 & 100 & [\citenum{tjornhammarfiles}]$^*$ & [\citenum{tjornhammar14}] \\
Botan-CHARMM36-UA~\cite{henin08} & DLPC & 128 & 3840 & 323 & 30 & 20 & [\citenum{charmmUAfiles}]        & SI \\
Lee-CHARMM36-UA~\cite{lee14}     & DPPC &  72 & 2189 & 323 & 70 & 50 & [\citenum{charmmUAfilesLEE}]$^*$ & SI \\
\end{tabular}
\end{sidewaystable}

\begin{table}[htb]
\centering
\caption{Simulated single component lipid bilayers with varying hydration levels. The simulation file data sets marked with $^*$ include also part of the trajectory.
$^a$ Water/lipid molar ratio
$^b$ The number of lipid molecules
$^c$ The number of water molecules
$^d$ Simulation temperature
$^e$ The total simulation time
$^f$ Time frames used in the analysis
$^g$ Reference link for the downloadable simulation files
$^h$ Reference for the full simulation details
}\label{systemsDEHYD}
\begin{tabular}{c c c c c c c c c c}
%\hline
Force field & lipid & $^a$n (w/l)   & $^b$N$_{\rm l}$   &  $^c$N$_{\rm w}$ & $^d$T (K)  & $^e$t$_{{\rm sim}}$(ns)  & $^f$t$_{{\rm anal}}$ (ns)& $^g$Files  &  $^h$Details\\
\hline
Berger-POPC-07~\cite{ollila07a}          &   POPC & 57  &128 & 7290  & 298  & 270 & 240 & [\citenum{bergerFILESpopc}]$^*$ & SI \\
                                        &   POPC & 7  &128 & 896   & 298  & 60 & 50 & [\citenum{bergerDEHYDfiles}]$^*$ & SI \\
Berger-DLPC-13~\cite{kanduc13}          &   DLPC & 28  &72 & 2016  & 300  & 80 & 60 & [\citenum{bergerFILESdlpc28}]$^*$ & [\citenum{kanduc13}] \\
                                 &   DLPC & 24  &72 & 1728  & 300  & 80 & 60 & [\citenum{bergerFILESdlpc24}]$^*$ & [\citenum{kanduc13}] \\
                                 &   DLPC & 20  &72 & 1440  & 300  & 80 & 60 & [\citenum{bergerFILESdlpc20}]$^*$ & [\citenum{kanduc13}] \\
                                 &   DLPC & 16  &72 & 1152  & 300  & 80 & 60 & [\citenum{bergerFILESdlpc16}]$^*$ & [\citenum{kanduc13}] \\
                                 &   DLPC & 12  &72 & 864  & 300  & 80 & 60 & [\citenum{bergerFILESdlpc12}]$^*$ & [\citenum{kanduc13}] \\
                                 &   DLPC & 8  &72 & 576  & 300  & 80 & 60 & [\citenum{bergerFILESdlpc8}]$^*$ & [\citenum{kanduc13}] \\
                                 &   DLPC & 4  &72 & 288  & 300  & 80 & 60 & [\citenum{bergerFILESdlpc4}]$^*$ & [\citenum{kanduc13}] \\
CHARMM36\cite{klauda10}          & POPC   & 40 & 128 &  5120   & 303 & 150 & 100  & [\citenum{charmm36files}]$^*$   & SI \\
                                 & POPC   & 31  & 72  &  2242 & 303 & 30 & 20 & [\citenum{charmm36filesSHORT}]$^*$ & SI \\
                               & POPC   & 15 & 72 &  1080  & 303 & 59 & 40 & [\citenum{charmm36files15wPERl}]$^*$ & SI \\
                            & POPC   & 7  & 72  &  504  & 303 & 60 & 20 & [\citenum{charmm36files7wPERl}]$^*$ & SI \\
MacRog\cite{kulig15b}     & POPC   & 50 & 288  & 14400 & 310 & 90 & 40 & [\citenum{macrogdehydFILES}]$^*$ & SI \\
                               & POPC   & 25 & 288  & 7200 & 310 & 100 & 50 & [\citenum{macrogdehydFILES}]$^*$ & SI \\
                                & POPC   & 20 & 288  & 5760 & 310 & 100 & 50 & [\citenum{macrogdehydFILES}]$^*$ & SI \\
                                & POPC   & 15 & 288  & 4320 & 310 & 100 & 50 & [\citenum{macrogdehydFILES}]$^*$ & SI \\
                                & POPC   & 10 & 288  & 2880 & 310 & 100 & 50 & [\citenum{macrogdehydFILES}]$^*$ & SI \\
                                & POPC   & 5 & 288   & 1440 & 310 & 100 & 50 & [\citenum{macrogdehydFILES}]$^*$ & SI \\
GAFFlipid\cite{dickson12}      & POPC   & 31& 126  & 3948  & 303 & 137 & 32 & [\citenum{GAFFlipidFILES}]$^*$ & SI  \\
                               & POPC   & 7& 126  & 896   & 303 & 130 & 40 & [\citenum{gaffDEHYDfiles}]$^*$ & SI \\
\end{tabular}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[An \textsf{achemso} demo]
  {SUPPLEMENTARY INFORMATION: Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\abbreviations{IR,NMR,UV}
%\keywords{American Chemical Society, \LaTeX}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The "tocentry" environment can be used to create an entry for the
%% graphical table of contents. It is given here as some journals
%% require that it is printed as part of the abstract page. It will
%% be automatically moved as appropriate.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\begin{tocentry}

%Some journals require a graphical entry for the Table of Contents.
%This should be laid out ``print ready'' so that the sizing of the
%text is correct.

%Inside the \texttt{tocentry} environment, the font used is Helvetica
%8\,pt, as required by \emph{Journal of the American Chemical
%Society}.

%The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
%permitted for  \emph{Journal of the American Chemical Society}
%graphical table of content entries. The box will not resize if the
%content is too big: instead it will overflow the edge of the box.

%This box and the associated title will always be printed on a
%separate page at the end of the document.

%\end{tocentry}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%\begin{center}
%{\bf SUPPLEMENTARY INFORMATION}
%\end{center}
\section{Simulation details} 
\subsection{Berger based models}
For the Berger based models we use here the following naming convention: 
Berger - \{{\it molecule name}\} - \{{\it year when model published first time}\} \{{\it citation}\}.
The reason is that there are several different molecular topologies which are using the non-bonded parameters originally
developed by Berger et al.~\cite{berger97}. Thus the common factor in the Berger based models are the non-bonded parameters,
while the molecule specific parameters might somewhat vary. However, the majority of the molecular level topologies are 
relying (especially for the glycerol backbone and headgroup) on the parameters originally introduced by Marrink et al.~\cite{marrink98}.
This is the case for all the Berger based simulations discussed in this work.
